---
title: "Metatranscriptomics abundance distributions"
author: 
  - name: Lan Huong Nguyen
    affiliation: Institute for Computational and Mathematical Engineering, Stanford University, CA 94305
keywords: microbiome, metatranscriptomics.
date: "`r BiocStyle::doc_date()`"
output: 
  BiocStyle::html_document2:
    toc_float: true
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Gene expression time course data analysis}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r style, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
rm(list = ls())
library("BiocStyle")
library("rmarkdown")
options(width = 200) 
knitr::opts_chunk$set(
  message = FALSE, error = FALSE,
  warning = FALSE, fig.width = 10, fig.height = 7,
  fig.path = "figs/", dev='png') 
```


In this notebook we examine the distributions of 
the abundance measures obtained for metatranscriptomics
reads aligned to RefSeq database.

```{r}
library("tidyverse")
library("readxl")
library("data.table")
library("gridExtra")

datadir <- "/scratch/PI/sph/resilience/metatranscriptomics/processed/"

theme_set(theme_bw())
theme_updates <- theme(text = element_text(size =20))
```

```{r, echo = FALSE}
sum_skip_na <- function(x) {sum(x, na.rm = TRUE)}

replace_na_with_zero <- function(x) {
  x[is.na(x)] <- 0
  return(x)
}
# metat <- metat[, (sample_cols) := lapply(.SD, replace_na_with_zero), .SDcols = sample_cols]


plt_trend <- function(df, x, y, contour = FALSE, 
                      max_npoints = 10^4, 
                      log_x = TRUE, log_y = FALSE){
  dfsub <- df
  if(nrow(df) > max_npoints ) {
    dfsub <- df[sample(1:nrow(df), max_npoints), ]
  }
  plt <- ggplot(df, aes_string(x, y)) 
  
  if(!contour) {
    plt <- plt +
      stat_density_2d(
        aes(fill = ..density..), 
        geom = "raster", contour = FALSE) +
      scale_fill_distiller(palette= "Spectral", direction=-1) +
      geom_point(data = dfsub, alpha = 0.1, size = 0.7)
  } else {
     plt <- plt +
      geom_point(data = dfsub, alpha = 0.1, size = 0.7) +
      geom_density_2d(aes(colour = ..level..)) +
      scale_color_distiller(palette= "Spectral", direction=-1) 
  }
  
  if (log_x){
    plt <- plt + 
      scale_x_continuous(expand = c(0, 0), trans = "log10")
  }
  if (log_y) {
    plt <- plt + 
      scale_y_continuous(expand = c(0, 0), trans = "log10")
  } else {
    plt <- plt + 
      scale_y_continuous(
        limits = c(NA, quantile(df[[y]], 0.95)),
        expand = c(0, 0)) 
  }
  return(plt)
}
```


# Loading data

```{r}
metat <- fread(file.path(datadir, "final_results/abund_refseq_len_ratio.csv"))
gene_info <- fread(file.path(datadir, "final_results/gene_len_refseq_raw.csv"))

dim(metat)
dim(gene_info)

gene_info2 <- metat[, .(GeneID, Organism, Function)]
identical(gene_info2[order(GeneID)], 
          gene_info[order(GeneID), !c("Length"), with = FALSE])
rm(gene_info2)
metat <- metat[, !c("Organism", "Function"), with = FALSE]
```



# Basic data summary

The abundance measure for a gene $i$ in measurement (sample) $j$ - `metat[i, j]` -
we have, is the total number (sum) of basepairs in all reads mapped to gene $i$,
in measurement $j$ divided by the length of gene $i$. We will refer to
this ratio as **abundance** from this point on.

Note that for each read, the number of basepairs aligned to a reference sequence
(gene) is usually much smaller than the length of that reference sequence.

## Sample depths

```{r}
sample_depth <- colSums(metat[, !c("GeneID"), with = FALSE], na.rm = TRUE)
summary(sample_depth)
```


```{r raw-sample-depths}
qplot(sample_depth, bins = 50)
```


## Obtain statistics on features

The gene lengths vary quite a bit:

```{r}
summary(gene_info$Length)
```

```{r}
plt <- qplot(gene_info$Length, bins = 50) 
grid.arrange(plt, plt+scale_x_log10())
```


In the summary below we see that even the sum of (the total number of) 
basepairs over all reads aligned to a gene is still much smaller than 
the length of the entire gene. The mean over samples for the ratio
between the sum of all bp aligned and the length of the gene is
usually smaller than 1.



```{r}
gene_info <- merge(
  gene_info,
  metat[,
        .(GeneID, 
          mean = rowMeans(.SD, na.rm = TRUE),
          sum = rowSums(.SD, na.rm = TRUE),
          prev = rowSums(.SD > 0, na.rm = TRUE)
          ),
        .SDcols = setdiff(colnames(metat), "GeneID")],
  by = "GeneID"
  )

summary(gene_info)
```


## Data Filtering

We discard all the samples for which the coverage was too small.

```{r}
sample_depth[sample_depth < 500]
```

```{r}
metat <- metat[, c("GeneID", names(sample_depth)[sample_depth > 500]), with = FALSE]
dim(metat)
```

Below we plot the prevalence and mean abundance of genes across all samples.

```{r}
plt1 <- qplot(gene_info$mean, bins = 50) +
  scale_x_log10()
plt2 <- qplot(gene_info$prev, bins = 50) 
grid.arrange(plt1, plt2)
```

# Gene abundance trends and distributions

The following shows that most of the genes
are present only in a under 10 (out 380) different samples.


```{r}
quantile(gene_info$prev, seq(0, 1, 0.05))
```


```{r}
sample_cols <- setdiff(colnames(metat), "GeneID")
metat_fltr <- metat[
  , (sample_cols) := lapply(.SD, replace_na_with_zero), 
   .SDcols = sample_cols]

# We filter out the genes which occur in less than 5% samples
frac <- 0.05
metat_fltr <- metat_fltr[
  rowSums(metat_fltr[, !c("GeneID"), with = FALSE] > 0) > frac*length(sample_cols)]
dim(metat_fltr)
```

Sample depth before filtering

```{r}
sample_depth <- colSums(metat[, !c("GeneID"), with = FALSE], na.rm = TRUE)
summary(sample_depth)
```
Sample depth after filtering

```{r}
sample_depth_fltr <- colSums(metat_fltr[, !c("GeneID"), with = FALSE], na.rm = TRUE)
summary(sample_depth_fltr)
```

```{r}
smp_data <- data.frame(
  File_Name = as.character(colnames(metat_fltr[, !c("GeneID")])),
  stringsAsFactors = FALSE
  ) %>%
  mutate(
    Meas_ID = sapply(File_Name, function(x) strsplit(x,  "_")[[1]][1]),
    Subject = sapply(File_Name, function(x) strsplit(x,  "_")[[1]][2]),
    Subject = gsub("sw", "", Subject)
  ) %>%
  tibble::column_to_rownames("File_Name")
```


```{r}
dds <- DESeq2::DESeqDataSetFromMatrix(
  floor(metat_fltr[, !c("GeneID")]), colData = smp_data, design = ~1)
dds <- DESeq2::estimateSizeFactors(dds, type = "poscounts")
```


```{r}
qplot(sample_depth_fltr, DESeq2::sizeFactors(dds))
```

```{r}
sample_cols <- setdiff(colnames(metat_fltr), "GeneID")
metat_fltr_asinh <- copy(metat_fltr)[
  , (sample_cols) := lapply(.SD, asinh), 
   .SDcols = sample_cols]

metat_fltr_log <- copy(metat_fltr)[
  , (sample_cols) := lapply(.SD, log1p),
   .SDcols = sample_cols]
```


```{r}
gene_stats <- metat_fltr[, 
  .(GeneID,
  Sum = rowSums(.SD),
  Prev = rowSums(.SD > 0),
  Mean  =  rowMeans(.SD),
  Median = apply(.SD, 1, median),
  Var = apply(.SD, 1, var),
  SD = apply(.SD, 1, sd)),
  .SDcols = sample_cols
]
summary(gene_stats)
```

```{r}
gene_stats_asinh<- metat_fltr_asinh[, 
  .(GeneID,
  Sum = rowSums(.SD),
  Prev = rowSums(.SD > 0),
  Mean  =  rowMeans(.SD),
  Median = apply(.SD, 1, median),
  Var = apply(.SD, 1, var),
  SD = apply(.SD, 1, sd)),
  .SDcols = sample_cols
]
summary(gene_stats_asinh)
```

```{r}
gene_stats_log <- metat_fltr_log[,
  .(GeneID,
  Sum = rowSums(.SD),
  Prev = rowSums(.SD > 0),
  Mean  =  rowMeans(.SD),
  Median = apply(.SD, 1, median),
  Var = apply(.SD, 1, var),
  SD = apply(.SD, 1, sd)),
  .SDcols = sample_cols
]
summary(gene_stats_log)
```


```{r gene-sums-raw}
qplot(gene_stats$Sum, bins = 50) + scale_x_log10() +
  ggtitle("Distribution of gene abundance sums across all samples")
```

```{r gene-sums-asinh}
qplot(gene_stats_asinh$Sum, bins = 50) + scale_x_log10() +
  ggtitle("Distribution of asinh transformed gene abundance sums across all samples")
```

```{r gene-sums-log1p}
qplot(gene_stats_log$Sum, bins = 50) + scale_x_log10() +
  ggtitle("Distribution of log1p transformed gene abundance sums across all samples")
```


```{r mean-var-raw}
plt_trend(gene_stats, "Mean", "Var", max_npoints = 10^5,
          contour = TRUE, log_y = TRUE) +
  geom_abline(color = "red", slope = 1, intercept = 0) +
  ggtitle("Gene abundance: Mean vs Variance") 
```

```{r mean-var-log}
plt_trend(gene_stats_log, "Mean", "Var", max_npoints = 10^5,
          contour = TRUE, log_y = TRUE) +
  geom_abline(color = "red", slope = 1, intercept = 0) +
  ggtitle("Log1p gene abundance: Mean vs Variance") 

```


```{r mean-var-asinh}
plt_trend(gene_stats_asinh, "Mean", "Var", max_npoints = 10^5,
          contour = TRUE, log_y = TRUE) +
  geom_abline(color = "red", slope = 1, intercept = 0) +
  ggtitle("Asinh gene abundance: Mean vs Variance") 

```

```{r mean-var-asinh-no-log}
plt_trend(gene_stats_asinh, "Mean", "Var", max_npoints = 10^5,
          contour = TRUE, log_y = TRUE, log_x = FALSE) +
  ggtitle("Asinh gene abundance: Mean vs Variance") + xlim(NA, 1.5)

```

```{r mean-prev-raw}
plt_trend(gene_stats, "Mean", "Prev", 
          max_npoints = 10^5, contour = TRUE, log_y = FALSE) +
  ggtitle("Gene abundance: Mean vs Prevalence") +
  scale_y_continuous(
    limits = c(NA, quantile(gene_stats[["Prev"]], 0.99)),
    expand = c(0, 0))
```

```{r mean-prev-asinh}
plt_trend(gene_stats_asinh, "Mean", "Prev", 
          max_npoints = 10^5, contour = TRUE, 
          log_x = FALSE) +
  ggtitle("Asinh gene abundance: Mean vs Prevalence") +
  xlim(NA, 1.5) +
  scale_y_continuous(
    limits = c(NA, quantile(gene_stats[["Prev"]], 0.99)),
    expand = c(0, 0)) 
```

# Aggregated data 

Now, we aggregate the gene abundance table into an organism
and function abundance table and save it for later analysis.


```{r}
orgtab <- merge(metat, gene_info[, .(GeneID, Organism)], by=c("GeneID"))
orgtab <- orgtab[, !c("GeneID")][, lapply(.SD, sum_skip_na), by=.(Organism)]
dim(orgtab)
```

```{r, eval = FALSE}
fwrite(orgtab, file.path(datadir, "final_results/organism_abundance_refseq.csv"))
```


It should be noted that genes from different organisms performing
the same functions with be merged together in the following table.


```{r}
funtab <- merge(metat, gene_info[, .(GeneID, Function)], by=c("GeneID"))
funtab <- funtab[, !c("GeneID")][, lapply(.SD, sum_skip_na), by=.(Function)]
dim(funtab)
```


```{r, eval = FALSE}
fwrite(funtab, file.path(datadir, "final_results/function_abundance_refseq.csv"))

rm(orgtab, funtab)
```


We can inspect the number of distinct organisms and functions
appearing in the study particinants data.

```{r}
gene_info$Organism <- gsub("\\[", "", gene_info$Organism)

cat("Number of different genes: ", nrow(gene_info), "\n")
cat("Number of unique Organisms: ", length(unique(gene_info$Organism)), "\n")
cat("Number of unique Functions: ", length(unique(gene_info$Function)), "\n")

```


Top 10 organisms with most abundant genes:

```{r}
gene_info[, .(total_sum = sum(sum)), by = .(Organism)][order(-total_sum)][1:10]
```

Top 10 organisms with most prevalent genes:

```{r}
gene_info[, .(total_prev = sum(prev)), by = .(Organism)][order(-total_prev)][1:10]
```



As expected the top functions are the ones with an uninformative 
label "hypothetical protein".


## Data aggregated by organism:


```{r, include=FALSE}
# Now you can restart an rsession and reload packages to clean up memory
rm(metat, sample_depth, gene_info)
library("data.table")
library("gridExtra")
library("tidyverse")
library("phyloseq")
library("readxl")

theme_set(theme_bw())
theme_updates <- theme(text = element_text(size =20))
```

```{r}
datadir <- "/scratch/PI/sph/resilience/metatranscriptomics/processed/"
orgtab <- fread(file.path(datadir, "final_results/organism_abundance_refseq.csv"))

```


```{r}
orgtab <- orgtab %>%
  mutate(
    Organism = gsub("'", "", Organism),
    Organism = gsub("\\[", "", Organism),
    Org = gsub("unclassified ", "", Organism),
    Org = gsub("uncultured ", "", Org),
    Org = gsub("unicellular ", "", Org)
    ) 

orgtab <- orgtab %>%
  separate(Org, c("Genus", "Species", "Strain"), 
           remove = TRUE, extra = "merge") 
```



```{r}
taxrank <- c("Organism", "Kingdom", "Phylum", "Class", 
            "Order", "Family", "Genus", "Species", "Strain")
sample_cols <- setdiff(colnames(orgtab), taxrank)
nsamples <- length(sample_cols)

orgtab <- data.table(orgtab)
orgtab_asinh <- copy(orgtab)[
  , (sample_cols) := lapply(.SD, asinh), 
   .SDcols = sample_cols]

org_stats <- orgtab[, 
  .(Organism,
    Sum = rowSums(.SD),
    Prev = rowSums(.SD > 0),
    Mean  =  rowMeans(.SD),
    Median = apply(.SD, 1, median),
    Var = apply(.SD, 1, var),
    SD = apply(.SD, 1, sd)),
  .SDcols = sample_cols
]

summary(org_stats)
```

```{r}
org_stats_asinh <- orgtab_asinh[, 
  .(Organism,
    Sum = rowSums(.SD),
    Prev = rowSums(.SD > 0),
    Mean  =  rowMeans(.SD),
    Median = apply(.SD, 1, median),
    Var = apply(.SD, 1, var),
    SD = apply(.SD, 1, sd)),
  .SDcols = sample_cols
]

summary(org_stats)
```


```{r org-sum}
qplot(org_stats$Sum, bins = 50) + scale_x_log10() +
  ggtitle("Organism sums across samples")
```

```{r org-sum}
qplot(org_stats_asinh$Sum, bins = 50) + scale_x_log10() +
  ggtitle("Asinh organism sums across samples")
```

```{r org-mean-var-raw}
plt_trend(org_stats, "Mean", "Var", contour = TRUE, log_y = TRUE) +
  ggtitle("Organism abundance: Mean vs Variance") +
  geom_abline(slope = 1, color = "red")
```

```{r org-mean-var-asinh}
plt_trend(org_stats_asinh, "Mean", "Var", contour = TRUE, log_y = TRUE) +
  geom_abline(slope = 1, color = "red") +
  ggtitle("Asinh organism abundance: Mean vs Variance") 
```


```{r org-mean-var-asinh-no-log}
plt_trend(org_stats_asinh, "Mean", "Var", contour = TRUE, 
          log_x = FALSE, log_y = TRUE) +
  ggtitle("Asinh organism abundance: Mean vs Variance") +
  xlim(NA, 1.5)
```

```{r org-mean-prev-raw}
plt_trend(org_stats, "Mean", "Prev", 
          max_npoints = 10^5, contour = TRUE) +
 ggtitle("Organism abundance: Mean vs Prevalence")
```

```{r org-mean-prev-asinh}
plt_trend(org_stats_asinh, "Mean", "Prev", 
          max_npoints = 10^5, contour = TRUE, log_x = FALSE) +
  ggtitle("Asinh organism abundance: Mean vs Prevalence") +
  xlim(NA, 1.5)
```

Now we are reading in 16s data for comparison and to obtain taxonomic
classification.

```{r}
ps <- readRDS(file.path(datadir, "../../16s/perturb_physeq_22Jan18.rds"))
```


```{r}
taxtab_16s <- data.table(as(ps@tax_table, "matrix"))[, 1:7]
taxtab_16s <- unique( taxtab_16s )
cat("Number of distinct species in 16s data:", nrow(taxtab_16s), "\n")
cat("Number of distinct organisms in metatranscriptomics data:", nrow(orgtab), "\n")
```

```{r, echo = FALSE}
add_ranks <- function(x, add_taxrank){
  for (t in 5:2) {
    ranktab <- add_taxrank[, 1:t]
    ranktab <- unique(ranktab)
    z <- data.table(
      matrix(NA,  
             nrow = nrow(ranktab),
             ncol = ncol(add_taxrank) - ncol(ranktab)
      )
    )
    ranktab <- cbind(ranktab, z)
    colnames(ranktab) <- colnames(add_taxrank)
    res <- merge(ranktab, x[, !c("Species"), with = FALSE],
                 by.x=colnames(ranktab)[t], by.y = "Genus")
    res <- res[, .(Organism, Kingdom, Phylum, Class, Order, Genus, Species)]
    if(nrow(res) != 0){ return(res) }
  }
  res <- data.table(Organism = x[, 1], Kingdom = NA, Phylum = NA, 
                      Class = NA, Order = NA, Genus = NA, Species = NA)
  return(res)
}

merge_any_level <- function(taxtab, add_taxrank){
  cols <- c("Organism", "Kingdom", "Phylum", "Class", 
            "Order", "Family", "Genus", "Species", "Strain")
  res <- merge(taxtab, add_taxrank, by = c("Genus", "Species"))
  res <- res[, cols, with = FALSE]
  remaining <- taxtab[!Organism %in% res$Organism]
  
  # 5:2 since we assigning from lowest rank "Genus" to "Phylum"
  ranktabs <- lapply(6:2, function(t){
    ranktab <- add_taxrank[, 1:t]
    ranktab <- unique(ranktab)
    ranktab <- ranktab[!is.na(ranktab[[t]])]
    z <- data.table(
      matrix(NA,  
             nrow = nrow(ranktab),
             ncol = ncol(add_taxrank) - ncol(ranktab) 
      )
    )
    ranktab <- cbind(ranktab, z)
    colnames(ranktab) <- colnames(add_taxrank)
    return(ranktab)
  })
  names(ranktabs) <- colnames(add_taxrank)[6:2]
  
  for (i in 1:length(ranktabs)) {
    rres <- merge(ranktabs[[i]],
                  remaining[, !c("Species"), with = FALSE], 
                  by.x = names(ranktabs)[i],
                  by.y = "Genus")
    rres <- rres[, cols, with = FALSE]
    res <- rbind(res, rres)
    remaining <- remaining[!Organism %in% res$Organism]
  }
  
  z <- data.table(
    matrix(NA, nrow(remaining), ncol(res) - ncol(remaining)))
  colnames(z) <- setdiff(colnames(res), colnames(remaining))
  remaining <- cbind(remaining, z)
  remaining <- remaining[, cols, with = FALSE]
  res <- rbind(res, remaining)
  res <- res[order(Organism)]
  return(res)
}
```

```{r}
orgnames <- c("Organism", "Genus", "Species", "Strain")
taxtab_metat <- data.table(orgtab)[, orgnames, with = FALSE]
taxtab_metat <- merge_any_level(taxtab_metat, taxtab_16s) 

```


```{r}
orgtab <- merge(taxtab_metat,
  data.table(orgtab)[, !c("Genus", "Species", "Strain"), with = FALSE],
  by = "Organism")

```


```{r}
genustab <- orgtab[,
   lapply(.SD, sum), 
   by = .(Kingdom, Phylum, Class, Order, Family, Genus), 
  .SDcols = sample_cols]
dim(genustab)
```

```{r}
taxranks <- c("Kingdom","Phylum", "Class", "Order", "Family",  "Genus")
metat_samples <- setdiff(colnames(genustab), taxranks)
genus_stat <- genustab[, .(Phylum, Class, Order, Family, Genus, sum = rowSums(.SD)), 
         .SDcols = metat_samples][order(-sum)]
genus_stat[1:10]
```

```{r}
set.seed(123456)
genera_ord <- genustab[
  , .(Phylum, Class, Order, Family, Genus, sum = rowSums(.SD)), 
  .SDcols = sample_cols][order(-sum)][!is.na(Genus)]

interval <- floor(seq(1, nrow(genera_ord), length.out = 6))
keep_idx <- c(sample(1:interval[2], 10), 
              sample(interval[3]:interval[4], 10),
              sample(interval[5]:nrow(genera_ord), 10))
example_genera <- genera_ord[keep_idx, Genus]


melt_example_genus <- genustab[Genus %in% example_genera] %>%
  gather(Meas_ID, abundance, -(Kingdom:Genus)) %>%
  mutate(Genus = factor(Genus, levels = example_genera))
```




```{r genera-hist-raw, fig.width=12, fig.height=10}
ggplot(melt_example_genus) +
  geom_histogram(
    aes(x = abundance, fill = Family),
    bins = 30) +
  facet_wrap(~ Genus, scale = "free", ncol = 5)
```


```{r genera-hist-asinh, fig.width=20, fig.height=12}
genustab_asinh <- copy(genustab)[
  , (sample_cols) := lapply(.SD, asinh), 
   .SDcols = sample_cols]

melt_example_genus_asinh <- genustab_asinh[Genus %in% genera_ord$Genus[1:20]] %>%
  gather(Meas_ID, abundance, -(Kingdom:Genus)) %>%
  mutate(Genus = factor(Genus, levels = genera_ord$Genus[1:20]))

ggplot(melt_example_genus_asinh) +
  geom_histogram(
    aes(x = abundance, fill = Family),
    bins = 30) +
  facet_wrap(~ Genus, scale = "free", ncol = 7) +
  theme(text = element_text(size = 20))
```

Now we will compare the metatranscriptomics abundance measures with 16s 
counts on a genus level.


First, we need to load the sample information file.

```{r}
meas <- read_xlsx(
  file.path(datadir, "../../sample_info/Mapping_Files_26March2018.xlsx"),
  "Meas", range = cell_cols(1:7))
names(meas) <- meas[1, ]
meas <- meas[-1, ] %>%
  rename(Samp_ID = SampID)

interv_levs <- c("NoInterv", "PreDiet", "MidDiet", "PostDiet", 
                 "PreCC", "MidCC","PostCC", "PreAbx", "MidAbx", "PostAbx")
samp <- read_xlsx(
  file.path(datadir, "../../sample_info/Mapping_Files_26March2018.xlsx"),
  "Samp", skip = 1) %>%
  mutate(
    Diet_Interval = ifelse(Diet_Interval == "NA", "NoInterv", Diet_Interval),
    CC_Interval = ifelse(CC_Interval == "NA", "NoInterv", CC_Interval),
    Abx_Interval = ifelse(Abx_Interval == "NA", "NoInterv", Abx_Interval),
    Diet_Interval = factor(Diet_Interval, interv_levs),
    CC_Interval = factor(CC_Interval, interv_levs),
    Abx_Interval = factor(Abx_Interval, interv_levs)
  ) %>%
  filter(Samp_Type != "ExtrCont")

meas <- meas %>% select(Meas_ID, Samp_ID) %>%
  left_join(samp)
rm(samp)
```

We convert the metatranscriptomics Genus abundance table
to long format.

```{r}
taxranks <- c("Kingdom","Phylum", "Class", "Order", "Family",  "Genus")
rel_total <- function(x) {x/sum(x)}

melt_genus_metat <- genustab[
  , (metat_samples) := lapply(.SD, rel_total), 
  .SDcols = metat_samples] %>%
  gather(Filename, rel_total, -(Kingdom:Genus)) %>%
  separate(Filename, c("Meas_ID", "Subject"), 
           remove = TRUE, extra = "drop") %>%
  left_join(meas %>% select("Meas_ID", "Samp_ID")) %>%
  mutate(Subject = gsub("sw", "", Subject)) %>%
  select(Kingdom:Genus, Meas_ID, Samp_ID, Subject, rel_total)

head(melt_genus_metat)
```


Then, we aggregate 16s data on Genus field:

```{r}
count16s <- as(ps@otu_table, "matrix") %>%
  data.frame() %>%
  rownames_to_column("Seq_id") %>%
  left_join(data.frame(ps@tax_table)[, 1:7] %>%
              rownames_to_column("Seq_id")) %>%
  data.table()

sample_cols <- colnames(count16s)[grepl("M", colnames(count16s))]
genus16s <- count16s[
  , lapply(.SD, sum), 
  by = .(Kingdom, Phylum, Class, Order, Family, Genus),
  .SDcols = sample_cols
][
  , (sample_cols) := lapply(.SD, rel_total), 
  .SDcols = sample_cols
]


melt_genus_16s <- genus16s %>%
  gather(Meas_ID, rel_total, -(Kingdom:Genus)) %>%
  left_join(meas %>% select("Meas_ID", "Samp_ID", "Subject")) %>%
  select(Kingdom:Genus, Meas_ID, Samp_ID, Subject, rel_total) 
head(melt_genus_16s)
```


```{r}

intersect_samples <- intersect(
  unique(melt_genus_metat$Samp_ID),
  unique(melt_genus_16s$Samp_ID))

length(intersect_samples) #375

melt_genus <- bind_rows(
  melt_genus_16s %>% 
    filter(Samp_ID %in% intersect_samples) %>%
    mutate(source = "16s"), 
  melt_genus_metat %>% 
    filter(Samp_ID %in% intersect_samples) %>%
    mutate(source = "metat")) 

## scatterplot of abundances from the two sources, per sample
genus_levels <- melt_genus %>%
  group_by(Genus) %>%
  summarise(mean = mean(rel_total)) %>%
  arrange(desc(mean))

melt_genus <- melt_genus  %>%
  mutate(Genus = factor(Genus, genus_levels$Genus))

subjects <- intersect(
  unique(melt_genus_16s$Subject),
  unique(melt_genus_metat$Subject))                    

scatter_data <- melt_genus %>%
  filter(Subject %in% subjects) %>%
  select(-Meas_ID) %>%
  spread(source, rel_total)

head(scatter_data)
```


```{r}
genus_subject_centroid <- scatter_data %>%
  group_by(Order, Family, Genus, Subject) %>%
  summarise(
    "16s" = median(`16s`),
    metat = median(metat, na.rm = TRUE)
  ) %>%
  as.data.frame()

```


```{r, echo = FALSE}
library(ggrepel)
plot_comparison <- function(scatter_data, genus_centroid,
                            x = "16s", y = "metatranscriptomic", 
                            color = "Diet_Interval", label = "Genus") {
  
  theme_scatter <- theme(
    legend.position = "bottom",  
    text = element_text(size = 15),
    strip.text = element_text(size = 15),
    strip.background = element_blank()
  )

  genus_centroid <- genus_centroid %>% as.data.frame() 
  scatter_data <- scatter_data %>% as.data.frame()
  
  genus_centroid$x <- genus_centroid[[x]]
  genus_centroid$y <- genus_centroid[[y]]
  scatter_data$x <- scatter_data[[x]]
  scatter_data$y <- scatter_data[[y]]
  
  genus_centroid <- genus_centroid %>%
    filter(!is.na(x) & !is.na(y))
  scatter_data <- scatter_data %>%
    filter(!is.na(x) & !is.na(y))
  
  data_repel <-  genus_centroid %>%
    filter(
      x > 0, y > 0, 
      abs(log10(x)- log10(y)) > 1.5
    )# %>% mutate(x = log10(x), y = log10(y) )
  
 #  genus_centroid[, c("x", "y")] <- log10(genus_centroid[, c("x", "y")])
 #  scatter_data[, c("x", "y")] <- log10(scatter_data[, c("x", "y")])
  
  ggplot() +
    geom_abline(alpha = 0.4, size = 0.2) +
    geom_point(
      data = scatter_data,
      aes_string("x", "y", col = color),
      size = 0.9, alpha = 0.5
    ) +
    geom_point(
      data = genus_centroid,
      aes_string(x = "x", y = "y", fill = color),
      size = 3, pch = 21, stroke = 0.2
    ) +
    geom_text_repel(
      data = data_repel,
      aes_string(x = "x", y = "y", label = label),
      size = 5,
      force = 0.2
    ) +
    facet_wrap(~Subject) + theme_scatter +
    scale_x_log10() + scale_y_log10() + 
    guides(color = "none") +
    xlab(x) + ylab(y) #xlim(log(-0.5), NA) 
}
```


```{r}
family_stat <- genus_stat %>%
  group_by(Phylum, Class, Order, Family) %>%
  summarise(sum = sum(sum)) %>%
  arrange(desc(sum))

tbl_df(family_stat)

top20_family <- family_stat %>% 
  filter(!is.na(Family)) 

top20_family <- top20_family[1:20, ] %>% .[["Family"]]

```


```{r metat-vs-16s, fig.width=8, fig.height=6}
subjects <- unique(scatter_data$Subject)
for (sub in subjects) {
  plt <- plot_comparison(
    scatter_data %>% 
      filter(Subject == sub) %>%
      mutate(Family = factor(Family, levels = top10_family)), 
    genus_subject_centroid %>% 
      filter(Subject == sub) %>%
      mutate(Family = factor(Family, levels = top10_family)),
    x = "16s", y = "metat",  
    color = "Family", label = "Genus")  
  print(plt)
}

```




## Data aggregated by function



```{r, include = FALSE}
funtab <- fread(file.path(datadir, "final_results/function_abundance_refseq.csv"))
```



```{r}
cat("Number of unique Functions: ", length(unique(gene_info$Function)), "\n")
```


Top 10 function with most abundant genes:

```{r}
gene_info[, .(total_sum = sum(sum)), by = .(Function)][order(-total_sum)][1:10]


```


Top 10 functions with most prevalent genes:

```{r}
gene_info[, .(total_prev = sum(prev)), by = .(Function)][order(-total_prev)][1:10]
```


```{r}
fun_stats <- funtab[, 
  .(Function,
    Sum = rowSums(.SD),
    Prev = rowSums(.SD > 0),
    Mean  =  rowMeans(.SD),
    Median = apply(.SD, 1, median),
    Var = apply(.SD, 1, var),
    SD = apply(.SD, 1, sd)),
  .SDcols = sample_cols
]

summary(fun_stats)
```

```{r}
sample_cols <- setdiff(colnames(funtab), "Function")
funtab_asinh <- copy(funtab)[
  , (sample_cols) := lapply(.SD, asinh), 
   .SDcols = sample_cols]

fun_stats_asinh <- funtab_asinh[, 
  .(Function,
    Sum = rowSums(.SD),
    Prev = rowSums(.SD > 0),
    Mean  =  rowMeans(.SD),
    Median = apply(.SD, 1, median),
    Var = apply(.SD, 1, var),
    SD = apply(.SD, 1, sd)),
  .SDcols = sample_cols
]

summary(fun_stats)
```


```{r fun-sum}
qplot(fun_stats$Sum, bins = 50) + scale_x_log10() +
  ggtitle("function sums across samples")
```

```{r fun-sum}
qplot(fun_stats_asinh$Sum, bins = 50) + scale_x_log10() +
  ggtitle("Asinh function sums across samples")
```

```{r fun-mean-var-raw}
plt_trend(fun_stats, "Mean", "Var", contour = TRUE, log_y = TRUE) +
  ggtitle("function abundance: Mean vs Variance") +
  geom_abline(slope = 1, color = "red")
```

```{r fun-mean-var-asinh}
plt_trend(fun_stats_asinh, "Mean", "Var", contour = TRUE, log_y = TRUE) +
  geom_abline(slope = 1, color = "red") +
  ggtitle("Asinh function abundance: Mean vs Variance") 
```


```{r fun-mean-var-asinh-no-log}
plt_trend(fun_stats_asinh, "Mean", "Var", contour = TRUE, 
          log_x = FALSE, log_y = TRUE) +
  ggtitle("Asinh function abundance: Mean vs Variance") +
  xlim(NA, 1.5)
```

```{r fun-mean-prev-raw}
plt_trend(fun_stats, "Mean", "Prev", 
          max_npoints = 10^5, contour = TRUE) +
 ggtitle("function abundance: Mean vs Prevalence")
```

```{r fun-mean-prev-asinh}
plt_trend(fun_stats_asinh, "Mean", "Prev", 
          max_npoints = 10^5, contour = TRUE, log_x = FALSE) +
  ggtitle("Asinh function abundance: Mean vs Prevalence") +
  xlim(NA, 1.5)
```

```{r}
head(funtab)
```

```{r}
top_functions
```


```{r}
top_20_function <- gene_info[, .(total_sum = sum(sum)), by = .(Function)][order(-total_sum)][1:20][["Function"]]
melt_example_function_asinh <- funtab_asinh[Function %in% top_20_function] %>%
  gather(Meas_ID, abundance, -Function) %>%
  mutate(Function = factor(Function, levels = top_20_function))
```



```{r function-hist-asinh, fig.width=30, fig.height=20}
ggplot(melt_example_function_asinh) +
  geom_histogram(
    aes(x = abundance),
    bins = 30) +
  facet_wrap(~ Function, scale = "free", ncol = 5) +
  theme(text = element_text(size = 25))
```

```{r}
top_20_function <- gene_info[, .(total_sum = sum(sum)), by = .(Function)]
top_20_function <- top_20_function[
  sapply(top_20_function$Function, function(x) !grepl("hypothetical", x))
]
top_20_function <- top_20_function[
  sapply(top_20_function$Function, function(x) !grepl("ribosomal", x))
]

top_20_function <- top_20_function[order(-total_sum)][1:20][["Function"]]
melt_example_function_asinh <- funtab_asinh[Function %in% top_20_function] %>%
  gather(Meas_ID, abundance, -Function) %>%
  mutate(Function = factor(Function, levels = top_20_function))
```



```{r function-hist-asinh-fltr, fig.width=30, fig.height=20}
ggplot(melt_example_function_asinh) +
  geom_histogram(
    aes(x = abundance),
    bins = 30) +
  facet_wrap(~ Function, scale = "free", ncol = 5) +
  theme(text = element_text(size = 25))
```

```{r}
data.frame(top20 = top_20_function)
```

# Ordinations


## Organism abundaces

```{r}
X <- t(as(orgtab[, !taxrank, with = FALSE], "matrix"))
smp_data <- meas %>%
  filter(Meas_ID %in% gsub("\\_(.*)", "", rownames(X))) %>%
  select(-contains("Date"), -Full_Code) %>%
  left_join(
    ps@sam_data %>% 
      select(Group:BirthYear, -contains("RelDay"), -contains("Interval")) %>%
      rename("Samp_ID" = SampID)
  )
```

```{r}
bray_dist <- vegan::vegdist(X, "bray")
```


```{r}
set.seed(123456)
tsne_bray <- Rtsne::Rtsne(
  bray_dist, dims = 2, 
  theta = 0.0,
  is_distance = TRUE, pca = FALSE,
  perplexity = 20, #min(30, floor((nrow(X) - 1)/3)),
  eta = 10, exaggeration_factor = 20)
  #eta = 10, exaggeration_factor = 12)
scores <- as.data.frame(tsne_bray$Y)
rownames(scores) <- rownames(X)
```


```{r}
colnames(scores) <- paste0("Axis", 1:ncol(scores))
scores_df <- scores %>%
  rownames_to_column("Meas_ID") %>%
  mutate(Meas_ID = gsub("\\_(.*)", "", Meas_ID)) %>%
  left_join(smp_data) %>%
  arrange(Group, Subject, Samp_Date)
```

```{r tsne-subj-org-metat, fig.width=8, fig.height=6}
(plt <- ggplot(scores_df, aes(Axis1, Axis2)) +
  geom_text(
    aes(label = Subject, color = Subject), size = 5) +
   scale_color_brewer(palette = "Set1") +
   ggtitle("tSNE plot on organism aggragated abundances") +
   theme(text = element_text(size = 15))
)
```


```{r tsne-subj-org-metat-abx, fig.width=8, fig.height=6}
library(viridis)
(plt <- ggplot(
  scores_df %>%
    mutate(Abx_RelDay = as.numeric(Abx_RelDay)),
  aes(Axis1, Axis2)
  ) +
  geom_text(
    aes(label = Subject, color = Abx_RelDay), size = 5) +
   scale_color_viridis(option = "inferno") +
   ggtitle("tSNE plot on organism aggragated abundances") +
   theme(text = element_text(size = 15), legend.position = "bottom"))
```


```{r tsne-subj-org-metat-time-diet, fig.width=8, fig.height=6}
(plt <- ggplot(
  scores_df %>%
    mutate(Diet_RelDay = as.numeric(Diet_RelDay)),
  aes(Axis1, Axis2)
  ) +
  geom_text(
    aes(label = Subject, color = Diet_RelDay), size = 5) +
   scale_color_viridis() +
   ggtitle("tSNE plot on organism aggragated abundances") +
   theme(text = element_text(size = 15), legend.position = "bottom")) 
```

```{r}
X_asinh <- t(as(orgtab_asinh[, !taxrank, with = FALSE], "matrix"))

pca_res <- prcomp(scale(X_asinh, scale = FALSE))
scores_pca <- as.data.frame(pca_res$x)
loadings_pca <- as.data.frame(pca_res$rotation)
eigs_pca <- data.frame(
  eig_idx = 1:length(pca_res$sdev), 
  eig = pca_res$sdev^2) %>%
  mutate(var_exp = 100*eig/sum(eig))
```


```{r}
colnames(scores_pca) <- paste0("Axis", 1:ncol(scores_pca))
scores_pca_df <- scores_pca %>%
  rownames_to_column("Meas_ID") %>%
  mutate(Meas_ID = gsub("\\_(.*)", "", Meas_ID)) %>%
  left_join(smp_data) %>%
  arrange(Group, Subject, Samp_Date)
```

```{r pca-subj-org-metat, fig.width=8, fig.height=6}
var.explained <- round(eigs_pca$var_exp, 2)
(plt <- ggplot(scores_pca_df, aes(Axis1, Axis2)) +
   geom_text(
    aes(label = Subject, color = Subject), 
    size = 5) +
   scale_color_brewer(palette = "Set2") +
   ggtitle("PCA plot on organism aggragated abundances") +
   theme(text = element_text(size = 15))  + 
  labs(
    x = sprintf("Axis1 [%s%% variance]", var.explained[1]),
    y = sprintf("Axis2 [%s%% variance]", var.explained[2])
  ) +
  coord_fixed(max(1, sqrt(var.explained[2] / var.explained[1])))
)
```

```{r pca-subj-org-metat-diet, fig.width=8, fig.height=6}
var.explained <- round(eigs_pca$var_exp, 2)
(plt <- ggplot(
  scores_pca_df %>%
    mutate(Diet_RelDay = as.numeric(Diet_RelDay)), aes(Axis1, Axis2)) +
   geom_text(
    aes(label = Subject, color = Diet_RelDay), 
    size = 5) +
   scale_color_viridis() +
   ggtitle("PCA plot on organism aggragated abundances") +
   theme(text = element_text(size = 15))  + 
  labs(
    x = sprintf("Axis1 [%s%% variance]", var.explained[1]),
    y = sprintf("Axis2 [%s%% variance]", var.explained[2])
  ) +
  coord_fixed(max(1, sqrt(var.explained[2] / var.explained[1])))
)
```

```{r pca-subj-org-metat-abx, fig.width=8, fig.height=6}
var.explained <- round(eigs_pca$var_exp, 2)
(plt <- ggplot(
  scores_pca_df %>%
    mutate(Abx_RelDay = as.numeric(Abx_RelDay)), aes(Axis1, Axis2)) +
   geom_text(
    aes(label = Subject, color = Abx_RelDay), 
    size = 5) +
   scale_color_viridis(option = "magma") +
   ggtitle("PCA plot on organism aggragated abundances") +
   theme(text = element_text(size = 15))  + 
  labs(
    x = sprintf("Axis1 [%s%% variance]", var.explained[1]),
    y = sprintf("Axis2 [%s%% variance]", var.explained[2])
  ) +
  coord_fixed(max(1, sqrt(var.explained[2] / var.explained[1])))
)
```

```{r}
plot_scores_time <- function(scores, size = 3, text = FALSE, 
                             eigs = NULL, path = FALSE){
  scores <- scores %>% 
    arrange(Subject, DayFromStart)
  plt <- ggplot(
    scores %>% filter(Subject != "NA_QC"), 
    aes_string(x = "Axis1", y = "Axis2")
  ) 
  if(path) {
    plt <- plt + 
      geom_path(
        aes(group = Subject, color = DayFromStart), 
        alpha = 0.8, lwd = 1) 
  }
  if(text){
    plt <- plt + 
      geom_text(aes(color = DayFromStart, label = Subject))
    
  } else {
    plt <- plt +
      geom_point(aes(color = DayFromStart), size = size)  
  }
  plt <- plt +
    geom_point(
      data = scores %>% filter(Timeline != "typical"), 
      aes(fill = Timeline, shape = Timeline),
      size = 2*size, lwd=10
    ) +
    scale_color_viridis() + 
    scale_shape_manual(values = c(23:25, 21, 22)) 

  if(!is.null(eigs)){
    var.explained <- round(100 * eigs/sum(eigs), 2)
    plt <- plt + 
      labs(
        x = sprintf("Axis1 [%s%% variance]", var.explained[1]),
        y = sprintf("Axis2 [%s%% variance]", var.explained[2])
      ) +
      coord_fixed(sqrt(var.explained[2] / var.explained[1])) 
  }
  return(plt + theme(text = element_text(size = 20))) 
} 
```


```{r fig.width=12, fig.height=5}
plot_scores_time(scores_df , path = TRUE) +
  facet_wrap(~Subject, scales = "free")
```



## Function

## Organism abundaces

```{r}
X <- t(as(funtab[, !c("Function"), with = FALSE], "matrix"))
X_asinh <- t(as(funtab_asinh[, !c("Function"), with = FALSE], "matrix"))

bray_dist <- vegan::vegdist(X, "bray")
```

```{r}
set.seed(123456)
tsne_bray <- Rtsne::Rtsne(
  bray_dist, dims = 2, 
  theta = 0.0,
  is_distance = TRUE, pca = FALSE,
  perplexity = 30, #min(30, floor((nrow(X) - 1)/3)),
  eta = 100, exaggeration_factor = 12)
  #eta = 10, exaggeration_factor = 12)
scores <- as.data.frame(tsne_bray$Y)
rownames(scores) <- rownames(X)
```


```{r}
set.seed(123456)
tsne_asinh <- Rtsne::Rtsne(
  X_asinh, dims = 2, 
  theta = 0.0,
  is_distance = FALSE, pca = TRUE,
  perplexity = 30)
# , #min(30, floor((nrow(X) - 1)/3)),
#   eta = 100, exaggeration_factor = 12)
#   #eta = 10, exaggeration_factor = 12)
scores <- as.data.frame(tsne_asinh$Y)
rownames(scores) <- rownames(X)
```



```{r}
colnames(scores) <- paste0("Axis", 1:ncol(scores))
scores_df <- scores %>%
  rownames_to_column("Meas_ID") %>%
  mutate(Meas_ID = gsub("\\_(.*)", "", Meas_ID)) %>%
  left_join(smp_data) %>%
  arrange(Group, Subject, Samp_Date)
```

```{r tsne-subj-fun-metat, fig.width=8 fig.height=6}
(plt <- ggplot(scores_df, aes(Axis1, Axis2)) +
  geom_text(
    aes(label = Subject, color = Subject), size = 5) +
   scale_color_brewer(palette = "Set1") +
   ggtitle("tSNE plot on function aggragated abundances") +
   theme(text = element_text(size = 15))
)
```


```{r tsne-subj-fun-metat-abx, fig.width=8, fig.height=6}
library(viridis)
(plt <- ggplot(
  scores_df %>%
    mutate(Abx_RelDay = as.numeric(Abx_RelDay)),
  aes(Axis1, Axis2)
  ) +
  geom_text(
    aes(label = Subject, color = Abx_RelDay), size = 5) +
   scale_color_viridis(option = "inferno") +
   ggtitle("tSNE plot on organism aggragated abundances") +
   theme(text = element_text(size = 15), legend.position = "bottom"))
```


```{r tsne-subj-org-metat-time-diet, fig.width=8, fig.height=6}
(plt <- ggplot(
  scores_df %>%
    mutate(Diet_RelDay = as.numeric(Diet_RelDay)),
  aes(Axis1, Axis2)
  ) +
  geom_text(
    aes(label = Subject, color = Diet_RelDay), size = 5) +
   scale_color_viridis() +
   ggtitle("tSNE plot on organism aggragated abundances") +
   theme(text = element_text(size = 15), legend.position = "bottom")) 
```


```{r fig.width=12, fig.height=5}
plot_scores_time(scores_df , path = TRUE) +
  facet_wrap(~Subject, scales = "free")
```

```{r}

pca_res <- prcomp(scale(X_asinh, scale = FALSE))
scores_pca <- as.data.frame(pca_res$x)
loadings_pca <- as.data.frame(pca_res$rotation)
eigs_pca <- data.frame(
  eig_idx = 1:length(pca_res$sdev), 
  eig = pca_res$sdev^2) %>%
  mutate(var_exp = 100*eig/sum(eig))
```


```{r}
colnames(scores_pca) <- paste0("Axis", 1:ncol(scores_pca))
scores_pca_df <- scores_pca %>%
  rownames_to_column("Meas_ID") %>%
  mutate(Meas_ID = gsub("\\_(.*)", "", Meas_ID)) %>%
  left_join(smp_data) %>%
  arrange(Group, Subject, Samp_Date)
```

```{r pca-subj-fun-metat, fig.width=8, fig.height=6}
var.explained <- round(eigs_pca$var_exp, 2)
(plt <- ggplot(scores_pca_df, aes(Axis1, Axis2)) +
   geom_text(
    aes(label = Subject, color = Subject), 
    size = 5) +
   scale_color_brewer(palette = "Set1") +
   ggtitle("PCA plot on organism aggragated abundances") +
   theme(text = element_text(size = 15))  + 
  labs(
    x = sprintf("Axis1 [%s%% variance]", var.explained[1]),
    y = sprintf("Axis2 [%s%% variance]", var.explained[2])
  ) +
  coord_fixed(max(1, sqrt(var.explained[2] / var.explained[1])))
)
```

```{r pca-subj-fun-metat-diet, fig.width=8, fig.height=6}
var.explained <- round(eigs_pca$var_exp, 2)
(plt <- ggplot(
  scores_pca_df %>%
    mutate(Diet_RelDay = as.numeric(Diet_RelDay)), aes(Axis1, Axis2)) +
   geom_text(
    aes(label = Subject, color = Diet_RelDay), 
    size = 5) +
   scale_color_viridis() +
   ggtitle("PCA plot on organism aggragated abundances") +
   theme(text = element_text(size = 15))  + 
  labs(
    x = sprintf("Axis1 [%s%% variance]", var.explained[1]),
    y = sprintf("Axis2 [%s%% variance]", var.explained[2])
  ) +
  coord_fixed(max(1, sqrt(var.explained[2] / var.explained[1])))
)
```

```{r pca-subj-fun-metat-abx, fig.width=8, fig.height=6}
var.explained <- round(eigs_pca$var_exp, 2)
(plt <- ggplot(
  scores_pca_df %>%
    mutate(Abx_RelDay = as.numeric(Abx_RelDay)), aes(Axis1, Axis2)) +
   geom_text(
    aes(label = Subject, color = Abx_RelDay), 
    size = 5) +
   scale_color_viridis(option = "magma") +
   ggtitle("PCA plot on organism aggragated abundances") +
   theme(text = element_text(size = 15))  + 
  labs(
    x = sprintf("Axis1 [%s%% variance]", var.explained[1]),
    y = sprintf("Axis2 [%s%% variance]", var.explained[2])
  ) +
  coord_fixed(max(1, sqrt(var.explained[2] / var.explained[1])))
)
```




# Read coverage 


```{r}
stats <- read.csv("/scratch/PI/sph/resilience/metatranscriptomics/stats/metat_workflow_stats.csv")
head(stats)
```

```{r}
stats <- stats %>%
  mutate(
    Filename = as.character(Meas_ID),
    Subject = sapply(Filename, function(x) strsplit(x, "_")[[1]][2]), 
    Meas_ID = sapply(Filename, function(x) strsplit(x, "_")[[1]][1]),
    Subject = sapply(Subject, function(x) gsub("sw", "", x)) 
  )
dim(stats)
```

```{r}
stats <- stats %>%
  filter(!is.na(input_fwd))
dim(stats)
head(stats)
```


```{r fig.width=12, fig.height=8}
preproc_cols <- 
  c("input_fwd", "trimmed", "qual_fltr", 
    "mRNA", "dmnd_refseq", "dmnd_seed")

melt_preproc <- stats[, c("Meas_ID", "Subject", preproc_cols)] %>%
  gather(step, count, input_fwd:dmnd_seed) %>%
  mutate(step = factor(step, levels = preproc_cols))


ggplot(melt_preproc, aes(x = step, y=count)) +
  geom_violin() + 
  geom_jitter(
    aes(color=Subject),
    alpha = 0.5, height = 0, width = 0.1, size = 1)  +
  theme(
    text = element_text(size = 20),
    axis.text.x = element_text(angle = 90, size = 20))
```

```{r}
summary(stats[, preproc_cols]/1e6)
```


```{r}
sum_stats <- melt_preproc %>%
  group_by(step) %>%
  summarise(
    min = min(count),
    Q1 = quantile(count, 0.25),
    mean = mean(count),
    median = median(count),
    Q3 = quantile(count, 0.75),
    max = max(count)
    ) %>%
  mutate_at(.funs = function(x) {round(x/1e6, 2)}, 
            .vars = c("min", "Q1", "mean", "median","Q3", "max")) %>%
  gather(stat, value, -step) %>% 
  spread(step, value)   

sum_stats <- sum_stats %>% 
  slice(match(c("min", "Q1", "mean", "median","Q3", "max"), stat))
sum_stats
```

```{r}
sum_stats_frac <- sum_stats %>% 
   mutate_at(.funs = function(x) {round(x/sum_stats$input_fwd, 2)}, 
            .vars = setdiff(colnames(sum_stats), "stat"))
sum_stats_frac
```


```{r}

unique_cols <- 
  c("unique", "vector_bwa", "vector_blat", "human_bwa", 
    "human_blat", "mRNA_unique", "contigs")

melt_preproc <- stats[, c("Meas_ID", "Subject", preproc_cols)] %>%
  gather(step, count, input_fwd:unassembled_reads) %>%
  mutate(step = factor(step, levels = preproc_cols))


ggplot(melt_preproc, aes(x = step, y=count)) +
  geom_violin() + 
  geom_jitter(
    aes(color=Subject),
    alpha = 0.5, height = 0, width = 0.1, size = 1)  +
  theme(axis.text.x = element_text(angle = 90, size = 20))
```



# Session Info

```{r}
sessionInfo()
```

