---
title: "Functional PCA"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("phyloseq")
library("RColorBrewer")
library("tidyverse")
library("fdapace")

datadir <- "../../data/"
curdir <- getwd()
theme_set(theme_bw())
theme_update(text = element_text(20))

theme_subplot <- theme(
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "transparent",colour = NA),
      plot.background = element_rect(fill = "transparent",colour = NA)
    )

abx_intv_cols <- c("PreAbx" = "grey60", "MidAbx" = "#E41A1C", 
                   "PostAbx" = "#00BFC4", "UnpAbx" = "Purple")
diet_intv_cols <- c("PreDiet" = "grey60", "MidDiet" = "#FD8D3C", 
                    "PostDiet" = "#4DAF4A")
cc_intv_cols <- c("PreCC" = "grey60", "PostCC" = "#7A0177") #"#AE017E")

intv_cols <- c(abx_intv_cols, diet_intv_cols, cc_intv_cols, "NoInterv" = "grey60")
```



# Load Data

```{r}
# File generated in /perturbation_16s/analysis/analysis_summer2019/generate_phyloseq.rmd
<<<<<<< HEAD
psSubj <- readRDS("../../data/16S/phyloseq/perturb_physeq_participants_decontam_15Jul19.rds")
=======
psSubj <- readRDS(
  "../../data/16S/phyloseq/perturb_physeq_participants_decontam_15Jul19.rds")
>>>>>>> 4d2dd8ae0a201b86a8ba480455a3a3966ad60b7b
psSubj
# otu_table()   OTU Table:         [ 2425 taxa and 4402 samples ]
# sample_data() Sample Data:       [ 4402 samples by 40 sample variables ]
SMP <- data.frame(sample_data(psSubj))
SUBJ <- SMP %>% select(Subject, Age:BirthYear) %>% distinct()
<<<<<<< HEAD
=======
TAXTAB <- data.frame(as(tax_table(psSubj), "matrix")) %>%
  rownames_to_column("Seq_ID") %>%
  select(1:7) 
>>>>>>> 4d2dd8ae0a201b86a8ba480455a3a3966ad60b7b
```


```{r}
load("output/pairwise_dist_to_baseline_subj_16S.rda")
<<<<<<< HEAD
=======
load("output/fpca_clust_res_abx.rda")
load("output/fpca_clust_res_diet.rda")
>>>>>>> 4d2dd8ae0a201b86a8ba480455a3a3966ad60b7b
load("output/fpca_res.rda")
ls()
```


# Functional PCA 


```{r}
<<<<<<< HEAD
=======
# (...) - a list options of options to fdapace()
fpca_wrapper <- function(df.long, time_column, value_column, replicate_column,
                    feat_column = NULL, parallel = FALSE, ncores = 1, 
                    num_nonzero = 10, thresh = 0, clust_min_num_replicate = 15,
                    cluster = FALSE, cmethod = "EMCluster", K = 2, 
                    fpca_optns = NULL, fclust_optns = NULL){
  require(dplyr)
  if (!time_column %in% colnames(df.long)) {
    stop("'df.long' does not have", time_column, "as time column.")
  }
  if (!value_column %in% colnames(df.long)) {
    stop("'df.long' does not have", value_column, "as value column.")
  }
  if (!replicate_column %in% colnames(df.long)) {
    stop("'df.long' does not have", replicate_column, "as replicate id column.")
  }
  if (all(!is.null(feat_column), !feat_column %in% colnames(df.long))) {
    stop("'df.long' does not have", feat_column, "as feature id column.")
  }
  if(!is.null(feat_column)){
    features <- unique(df.long[[feat_column]])
  } else {
    feat_column <- "Feature_ID"
    features <- df.long[[feat_column]] <- "Feat1"
  }
  if(parallel) {
    require(doParallel)
    ncores <- min(ncores, detectCores())
    cat("Running on ", ncores, " cores...\n")
    registerDoParallel(ncores)
    res <- foreach(feat=features) %dopar% {
      fit_fdapca(df.long, time_column, value_column, 
                 replicate_column, feat_column, feat, 
                 cluster = cluster, cmethod = cmethod, K = K, 
                 num_nonzero = num_nonzero, thresh = thresh,  
                 clust_min_num_replicate = clust_min_num_replicate, 
                 fpca_optns = fpca_optns, fclust_optns = fclust_optns)
    }
  } else {
    res <- lapply(features, function(feat) {
      cat("Processing feature: ", feat, "...\n")
      fit_fdapca(df.long, time_column, value_column, 
                 replicate_column, feat_column, feat, 
                 cluster = cluster, cmethod = cmethod, K = K, 
                 num_nonzero = num_nonzero, thresh = thresh,  
                 clust_min_num_replicate = clust_min_num_replicate, 
                 fpca_optns = fpca_optns, fclust_optns = fclust_optns)
    })
  }
  names(res) <- features
  return(res)
}


>>>>>>> 4d2dd8ae0a201b86a8ba480455a3a3966ad60b7b
fit_fdapca <- function(
  df, time_column, value_column, replicate_column,
  feat_column = NULL, feat = NULL,
  cluster = FALSE, cmethod = "EMCluster", K = 2, filter = TRUE,
  thresh = 0, num_nonzero = 10, clust_min_num_replicate = 10,
  fpca_optns = NULL, fclust_optns = NULL)
{
  if(!is.null(feat) & !is.null(feat_column)) {
    df <- df %>% 
      filter_at(.vars = feat_column, any_vars((.) == feat))
  }
  y_lst <- plyr::dlply(df, replicate_column, function(x) x[[value_column]])
  t_lst <- plyr::dlply(df, replicate_column, function(x) x[[time_column]])
  if (filter){
    idx <- sapply(y_lst, function(y){sum(abs(y) > thresh) >= num_nonzero})
    y_lst <- y_lst[idx]
    t_lst <- t_lst[idx]
  }  
  if(length(y_lst) == 0) return(NULL)
  feat_res <- NULL
  if(cluster & (length(y_lst) >= clust_min_num_replicate)) {
    feat_res <- try(fdapace::FClust(
      y_lst, t_lst, k = K, optnsFPCA = fpca_optns, optnsCS = fclust_optns))
  } else {
    if(is.null(fpca_optns)) {
      fpca_optns <- list()
    } 
    if (length(y_lst) < clust_min_num_replicate) {
      fpca_optns$dataType <- 'Sparse'
    }
    feat_res <- fdapace::FPCA(y_lst, t_lst, optns = fpca_optns)
    if(cluster){
      feat_res <- list("cluster" = rep(1, length(y_lst)), 
                       "fpca" = feat_res, "clusterObj" = NULL)
    }
  }
  feat_res
}

fitted_values_fpca <- function(obj, derOptns = list(p = 0)) 
{
  selectedK <- NULL; clust_df <- NULL
  if("cluster" %in% names(obj)){
    clust_df <- data.frame(
      "Replicate_ID" = names(obj$fpca$inputData$Ly),
      "Cluster" =  as.character(obj[["cluster"]]),
      stringsAsFactors = FALSE)
    obj <- obj[["fpca"]]
  }
  if(derOptns$p > 0) {
    selectedK <- fdapace::SelectK(obj)$K
    if(!is.finite(selectedK)) selectedK <- ncol(obj$xiEst)
  }  
  fit <- fdapace:::fitted.FPCA(obj, K = selectedK, derOptns = derOptns)
  rownames(fit) <- names(obj$inputData$Ly)
  colnames(fit) <- obj$workGrid
  fit <- reshape2::melt(fit, varnames = c("Replicate_ID", "time")) %>%
    mutate(Replicate_ID = as.character(Replicate_ID))
  if(!is.null(clust_df)){
    fit <- suppressMessages(fit  %>% left_join(clust_df))
  }
  return(fit)
}


fit_dist_to_baseline <- function(
  dist_to_baseline, time_column="RelDay", 
  value_column = "dist_to_baseline", replicate_column = "Subject") 
{
  nGrid <- length(
    seq(min(dist_to_baseline[[time_column]]),
        max(dist_to_baseline[[time_column]])))
  fpca_res <- fit_fdapca(
    dist_to_baseline, time_column, value_column, replicate_column,
    cluster = FALSE, filter = FALSE, fpca_optns = list(nRegGrid = nGrid))
  
  fitted_mean <- data.frame(time = fpca_res$workGrid, value = fpca_res$mu)
  fitted_response <- fitted_values_fpca(fpca_res) %>%
    mutate(Subject = Replicate_ID) %>%
    left_join(
      fitted_values_fpca(fpca_res, derOptns = list(p = 1)) %>%
        mutate(Subject = Replicate_ID) %>%
        rename("deriv" = value)
    )
  return(list(res = fpca_res, mean = fitted_mean, fitted = fitted_response))
}
```


```{r}
# this is because some subjects have multiple samples take the same day
bray_to_baseline_fltr <- bray_to_baseline %>% 
  group_by(perturbation, Subject, Group, Interval, RelDay) %>%
  summarise(n = n(), dist_to_baseline = mean(dist_to_baseline)) %>%
  ungroup()

bray_to_baseline_fltr %>% filter(n > 1)
```


## Antibiotics


```{r bray-abx}
bray_to_baseline_fltr %>% 
  filter(perturbation == "Abx") %>%
  filter(RelDay >= -50, RelDay <= 60) %>%
  mutate(Interval = factor(Interval, level = names(abx_intv_cols))) %>%
  ggplot(
    aes(x = RelDay, y = dist_to_baseline, 
        group = Subject, color = Interval)) +
  geom_line(aes(group = Subject), alpha = 0.7, lwd = 0.5) + 
  geom_point(alpha = 0.5, size = 1.2) + 
  scale_color_manual(values = abx_intv_cols) + 
  theme(legend.position = "bottom") + 
<<<<<<< HEAD
  guides(colour = guide_legend(override.aes = list(size=3))) 
# +
#   xlab("Days from initial antibiotic dose") +
#   ylab("Bray-Curtis distance to 7 pre-antibiotic samples") 
=======
  guides(colour = guide_legend(override.aes = list(size=3))) +
  xlab("Days from initial antibiotic dose") +
  ylab("Bray-Curtis distance to 7 pre-antibiotic samples") 
>>>>>>> 4d2dd8ae0a201b86a8ba480455a3a3966ad60b7b
```



```{r, eval = FALSE}
fpca.bray.abx <- fit_dist_to_baseline(
  bray_to_baseline_fltr %>% filter(
    perturbation == "Abx", RelDay >= -50, RelDay <= 60))

save(list = c("fpca.bray.abx"), file = "output/fpca_res.rda")
```

```{r fpca-bray-abx, fig.width=12, fig.height=6.5}
(pAbx <- bray_to_baseline_fltr %>% 
  filter(perturbation == "Abx", RelDay >= -50, RelDay <= 60) %>%
ggplot(aes(x = RelDay, y = dist_to_baseline)) +
  geom_line(
    data = fpca.bray.abx[["fitted"]],
    aes(group = Subject, x = time, y = value),
    alpha = 0.3, size = 0.7, color = "grey30") +
  geom_vline(xintercept = 0, lwd = 1, color = "orange") +
  geom_vline(xintercept = 4, lwd = 1, color = "orange") +
  geom_point(aes(color = Interval), size = 1.5, alpha = 0.7) +
  geom_line(
    data = fpca.bray.abx[["mean"]], aes(x = time, y = value),
    color = "navy", size = 2) +
  scale_color_manual(values = abx_intv_cols, name = "Interval") +
  scale_x_continuous(
    name = "Days from initial antibiotic dose",
    limits = c(NA, NA), breaks = seq(-50, 60, 10)) +
  scale_y_continuous(
    name = "Bray-Curtis distance to baseline", 
    limits = c(0.1, 0.85), breaks = seq(0.1, 0.80, 0.1)) +
  theme(text = element_text(size = 20)))
```


```{r fpca-deriv-bray-abx, fig.width=12, fig.height=6.5}
fpca.bray.abx[["fitted"]] <- fpca.bray.abx[["fitted"]] %>%
  mutate(
    Interval = ifelse(fpca.bray.abx[["fitted"]]$time < 0 , "PreAbx",
               ifelse(fpca.bray.abx[["fitted"]]$time >= 0 & fpca.bray.abx[["fitted"]]$time <= 4, "MidAbx",
                      "PostAbx")))

(pAbxDeriv <-  fpca.bray.abx[["fitted"]] %>%
  ggplot(aes(x = RelDay)) +
  geom_line(
    aes(group = Subject, x = time, y = deriv, color = Interval),
    alpha = 0.5, size = 0.7) +
  geom_vline(xintercept = 0, lwd = 1, color = "orange") +
  geom_vline(xintercept = 4, lwd = 1, color = "orange") +
  scale_color_manual(values = abx_intv_cols, name = "Interval") +
  scale_x_continuous(name = "",
    limits = c(NA, NA), breaks = seq(-50, 60, 10)) +
  ylab("Derivative"))
```


```{r fpca-bray-abx-labs, fig.width=12, fig.height=6.5}
(pAbxLab <- bray_to_baseline_fltr %>% 
  filter(perturbation == "Abx", RelDay >= -50, RelDay <= 60) %>%
ggplot(aes(x = RelDay, y = dist_to_baseline)) +
  geom_line(
    data = fpca.bray.abx[["fitted"]],
    aes(group = Subject, x = time, y = value),
    alpha = 0.3, size = 0.7 ) +
  geom_vline(xintercept = 0, lwd = 1, color = "orange") +
  geom_vline(xintercept = 4, lwd = 1, color = "orange") +
  geom_text(
      aes(label = Subject, color = Interval), size = 4, alpha = 0.7) +
  geom_line(
    data = fpca.bray.abx[["mean"]], aes(x = time, y = value),
    color = "navy", size = 2) +
  scale_color_manual(values = abx_intv_cols, name = "Interval") +
  scale_x_continuous(
    name = "Days from initial antibiotic dose",
    limits = c(NA, NA), breaks = seq(-50, 60, 10)) +
  scale_y_continuous(
    name = "Bray-Curtis distance to baseline", 
    limits = c(0.1, 0.85), breaks = seq(0.1, 0.80, 0.1)) +
  theme(text = element_text(size = 20)))

  # geom_point(
  #     data = abx_bray %>% filter(RelDay > StabTime),
  #     color = "orange", size = 2.5) +
```

```{r fpca-bray-abx-val-deriv, fig.width=12, fig.height=6.5}
<<<<<<< HEAD
vp = grid::viewport(width = 0.3, height = 0.32, x = 0.07, y =0.95, just = c("left", "top"))
=======
vp = grid::viewport(width = 0.3, height = 0.32,
                    x = 0.07, y =0.95, just = c("left", "top"))
>>>>>>> 4d2dd8ae0a201b86a8ba480455a3a3966ad60b7b
print(pAbx)
print(pAbxDeriv + theme_bw(base_size = 10) + theme_subplot, vp = vp)
```


### Cluster RSVs

```{r, eval = FALSE}
<<<<<<< HEAD
abx.rsv <- data.frame(asinh(as(otu_table(abx), "matrix"))) %>%
  rownames_to_column("Seq_ID") %>%
  gather(Meas_ID, abundance, -Seq_ID) %>%
  left_join(abx_SMP) %>%
  left_join(taxtab) %>%
  mutate(
    Abx_Interval = factor(
      Abx_Interval, levels = c("PreAbx", "MidAbx", "PostAbx"))) %>%
  filter(Abx_RelDay >= -40, Abx_RelDay <= 80) %>%   # limit the scope due to effect of other perturbations
  arrange(Seq_ID, Subject, Abx_RelDay)
```

```{r, eval = FALSE}
## Run using  fclust.R script

thresh <- 0; num_nonzero <- 10; min_no_subj <- 3
keep_rsv <- abx.rsv %>%
  filter(abundance > thresh) %>%
  group_by(Subject, Seq_ID) %>%
  summarise(Freq = n()) %>%   # No. of samples for each subject with non-zero count of a given rsv
  filter(Freq >= num_nonzero) %>%
  group_by(Seq_ID) %>%
  summarise(Freq = n()) %>% # For each RSV, # of subjects w/ num_nonzero samples with counts > thresh
  filter(Freq >= min_no_subj) %>%
  arrange(desc(Freq))

length(unique(keep_rsv$Seq_ID)) # = 861
=======
abxFac <- c("PreAbx", "MidAbx", "PostAbx")
abx <- subset_samples(psSubj, Abx_RelDay >= -50, Abx_RelDay <= 60)
abx <- subset_taxa(abx, taxa_sums(abx) > 0)
abx

(replicated <- data.frame(sample_data(abx)) %>% 
  group_by(Subject, Group, Interval, Abx_RelDay) %>%
  mutate(n = n()) %>%
  ungroup() %>% filter(n > 1))

```

```{r, eval = FALSE}
# Remove replicated samples
sample_sums(abx)[replicated$Meas_ID]

#  M2242  M2243  M7869  M7886 
# 109520  88834  74697  74873 

# we retain a replicate with higher sample depth:

abx <- subset_samples(abx, !Meas_ID %in% c("M2243", "M7869"))
abx
```


```{r, eval = FALSE}
abx.rsv <- data.frame(asinh(as(otu_table(abx), "matrix"))) %>%
  rownames_to_column("Seq_ID") %>%
  gather(Meas_ID, abundance, -Seq_ID) %>%
  left_join(SMP) %>%
  left_join(TAXTAB) %>%
  mutate(Abx_Interval = factor(Abx_Interval, levels = abxFac)) %>%
  arrange(Seq_ID, Subject, Abx_RelDay)


thresh <- 0; num_nonzero <- 10; 
min_no_subj <- 3

keep_rsv <- abx.rsv %>%
  filter(abundance > thresh) %>%
  group_by(Subject, Seq_ID) %>%
  summarise(Freq = n()) %>%   # No. of samples per subject w/ positive rsv count 
  filter(Freq >= num_nonzero) %>%
  group_by(Seq_ID) %>%
  summarise(Freq = n()) %>% # Number of subjects w/ num_nonzero samples with counts > thresh
  filter(Freq >= min_no_subj) %>%
  arrange(desc(Freq))

length(unique(keep_rsv$Seq_ID)) # = 921
>>>>>>> 4d2dd8ae0a201b86a8ba480455a3a3966ad60b7b

abx.rsv.subset <- abx.rsv %>%
  filter(Seq_ID %in% unique(keep_rsv$Seq_ID))

<<<<<<< HEAD
save(list = c("abx.rsv", "keep_rsv"), file = "results/abx_long.rda")
save(list = c("keep_rsv", "abx.rsv.subset"), file = "results/abx_long_subset.rda")
```


```{r}

abx_fclust <- fda_pca(
=======
#save(list = c("abx.rsv", "keep_rsv"), file = "output/abx_long.rda")
save(list = c("keep_rsv", "abx.rsv.subset"), 
     file = "output/fpca_clust_res_abx.rda")
```


```{r, eval = FALSE}
source("./fda_pca_funs.R")
abx_fclust <- fpca_wrapper(
>>>>>>> 4d2dd8ae0a201b86a8ba480455a3a3966ad60b7b
  abx.rsv.subset,
  time_column = "Abx_RelDay",
  value_column = "abundance",
  replicate_column = "Subject",
  feat_column = "Seq_ID",
  cluster = TRUE,
  clust_min_num_replicate = 15,
  fpca_optns = NULL, fclust_optns = NULL,
  parallel = TRUE, ncores = 16)
save(list = c("abx_fclust"),
    file = "results/abx_rsv_fclust.rda")
<<<<<<< HEAD

abx_fclust_mu <- get_fpca_means(abx_fclust)
abx_fclust_subjFit <- get_fpca_fits(abx_fclust)

save(list = c("abx_fclust", "abx_fclust_subjFit", "abx_fclust_mu"),
    file = "results/abx_rsv_fclust.rda")
```

=======
abx_fclust_mu <- get_fpca_means(abx_fclust)
abx_fclust_subjFit <- get_fpca_fits(abx_fclust)

save(list = c("keep_rsv", "abx.rsv.subset", 
              "abx_fclust", "abx_fclust_subjFit", "abx_fclust_mu"),
    file = "output/fpca_clust_res_abx.rda")
```

```{r}

```



>>>>>>> 4d2dd8ae0a201b86a8ba480455a3a3966ad60b7b
## Diet


```{r}
bray_to_baseline_fltr %>% 
  filter(perturbation == "Diet", RelDay >= -30, RelDay <= 30) %>%
  mutate(Interval = factor(Interval, level = names(diet_intv_cols))) %>%
  ggplot(
    aes(x = RelDay, y = dist_to_baseline, 
        group = Subject, color = Interval)) +
  geom_line(aes(group = Subject), alpha = 0.7, lwd = 0.5) + 
  geom_point(alpha = 0.5, size = 1.2) + 
  scale_color_manual(values = diet_intv_cols) + 
  theme(legend.position = "bottom") + 
  guides(colour = guide_legend(override.aes = list(size=3))) +
  xlab("Days from diet initiation") +
  ylab("Bray-Curtis distance to 7 pre-diet samples") 
```


```{r, eval = FALSE}
fpca.bray.diet30 <- fit_dist_to_baseline(
  bray_to_baseline_fltr %>% filter(perturbation == "Diet", RelDay >= -30, RelDay <= 30))

save(list = c("fpca.bray.abx", "fpca.bray.diet", "fpca.bray.diet30"), file = "output/fpca_res.rda")
```

```{r fpca-bray-diet, fig.width=12, fig.height=6.5}
(pDiet <- bray_to_baseline_fltr %>% 
  filter(perturbation == "Diet", RelDay >= -30, RelDay <= 30) %>%
ggplot(aes(x = RelDay, y = dist_to_baseline)) +
  geom_line(
    data = fpca.bray.diet30[["fitted"]],
    aes(group = Subject, x = time, y = value),
    alpha = 0.3, size = 0.7, color = "grey30") +
  geom_vline(xintercept = 0, lwd = 1, color = "orange") +
  geom_vline(xintercept = 4, lwd = 1, color = "orange") +
  geom_point(aes(color = Interval), size = 1.5, alpha = 0.7) +
  geom_line(
    data = fpca.bray.diet30[["mean"]], aes(x = time, y = value),
    color = "navy", size = 2) +
  scale_color_manual(values = diet_intv_cols, name = "Interval") +
  scale_x_continuous(
    name = "Days from initial antibiotic dose",
    limits = c(NA, NA), breaks = seq(-50, 60, 10)) +
  scale_y_continuous(
    name = "Bray-Curtis distance to baseline", 
    limits = c(NA, NA), breaks = seq(0.1, 0.80, 0.1)) +
  theme(text = element_text(size = 20))) 
```


```{r fpca-deriv-bray-diet, fig.width=12, fig.height=6.5}
fpca.bray.diet30[["fitted"]] <- fpca.bray.diet30[["fitted"]] %>%
  mutate(
    Interval = ifelse(fpca.bray.diet30[["fitted"]]$time < 0 , "PreDiet",
               ifelse(fpca.bray.diet30[["fitted"]]$time >= 0 & fpca.bray.diet30[["fitted"]]$time <= 4, "MidDiet",
                      "PostDiet")))

(pDietDeriv <-  fpca.bray.diet30[["fitted"]] %>%
  ggplot(aes(x = RelDay)) +
  geom_line(
    aes(group = Subject, x = time, y = deriv, color = Interval),
    alpha = 0.5, size = 0.7) +
  geom_vline(xintercept = 0, lwd = 1, color = "orange") +
  geom_vline(xintercept = 4, lwd = 1, color = "orange") +
  scale_color_manual(values = diet_intv_cols, name = "Interval") +
  scale_x_continuous(name = "",
    limits = c(NA, NA), breaks = seq(-50, 60, 10)) +
  ylab("Derivative"))
```


```{r fpca-bray-diet-labs, fig.width=12, fig.height=6.5}
(pDietLab <- bray_to_baseline_fltr %>% 
  filter(perturbation == "Diet", RelDay >= -50, RelDay <= 60) %>%
ggplot(aes(x = RelDay, y = dist_to_baseline)) +
  geom_line(
    data = fpca.bray.diet[["fitted"]],
    aes(group = Subject, x = time, y = value),
    alpha = 0.3, size = 0.7 ) +
  geom_vline(xintercept = 0, lwd = 1, color = "orange") +
  geom_vline(xintercept = 4, lwd = 1, color = "orange") +
  geom_text(
      aes(label = Subject, color = Interval), size = 4, alpha = 0.7) +
  geom_line(
    data = fpca.bray.diet[["mean"]], aes(x = time, y = value),
    color = "navy", size = 2) +
  scale_color_manual(values = diet_intv_cols, name = "Interval") +
  scale_x_continuous(
    name = "Days from initial antibiotic dose",
    limits = c(NA, NA), breaks = seq(-50, 60, 10)) +
  scale_y_continuous(
    name = "Bray-Curtis distance to baseline", 
    limits = c(0.1, 0.85), breaks = seq(0.1, 0.80, 0.1)) +
  theme(text = element_text(size = 20)))

```


```{r fpca-bray-diet-val-deriv, fig.width=12, fig.height=6.5}
vp = grid::viewport(width = 0.3, height = 0.32, x = 0.07, y =0.95, just = c("left", "top"))
print(pDiet)
print(pDietDeriv + theme_bw(base_size = 10) + theme_subplot, vp = vp)
```

<<<<<<< HEAD
## Colon cleanout

=======

### Cluster RSVs

```{r, eval = FALSE}
dietFac <- c("PreDiet", "MidDiet", "PostDiet")
diet <- subset_samples(psSubj, Diet_RelDay >= -30, Diet_RelDay <= 30)
diet <- subset_taxa(diet, taxa_sums(diet) > 0)
diet
diet.rsv <- data.frame(asinh(as(otu_table(diet), "matrix"))) %>%
  rownames_to_column("Seq_ID") %>%
  gather(Meas_ID, abundance, -Seq_ID) %>%
  left_join(SMP) %>%
  left_join(TAXTAB) %>%
  mutate(Diet_Interval = factor(Diet_Interval, levels = dietFac)) %>%
  arrange(Seq_ID, Subject, Diet_RelDay)
```

```{r, eval = FALSE}
## Run using  fclust.R script
thresh <- 0; num_nonzero <- 10; 
min_no_subj <- 3

keep_rsv <- diet.rsv %>%
  filter(abundance > thresh) %>%
  group_by(Subject, Seq_ID) %>%
  summarise(Freq = n()) %>%   # No. of samples per subject w/ positive rsv count 
  filter(Freq >= num_nonzero) %>%
  group_by(Seq_ID) %>%
  summarise(Freq = n()) %>% # Number of subjects w/ num_nonzero samples with counts > thresh
  filter(Freq >= min_no_subj) %>%
  arrange(desc(Freq))

length(unique(keep_rsv$Seq_ID)) # = 921

diet.rsv.subset <- diet.rsv %>%
  filter(Seq_ID %in% unique(keep_rsv$Seq_ID))

save(list = c("keep_rsv", "diet.rsv.subset"), 
     file = "output/fpca_clust_res_diet.rda")
```


```{r}
source("./fda_pca_funs.R")
diet_fclust <- fpca_wrapper(
  diet.rsv.subset,
  time_column = "Diet_RelDay",
  value_column = "abundance",
  replicate_column = "Subject",
  feat_column = "Seq_ID",
  cluster = TRUE,
  clust_min_num_replicate = 15,
  fpca_optns = NULL, fclust_optns = NULL,
  parallel = TRUE, ncores = 16)


diet_fclust_mu <- get_fpca_means(diet_fclust)
diet_fclust_subjFit <- get_fpca_fits(diet_fclust)

save(list = c("keep_rsv", "diet.rsv.subset", 
              "diet_fclust", "diet_fclust_subjFit", "diet_fclust_mu"),
    file = "output/fpca_clust_res_diet.rda")
```

## Colon cleanout
>>>>>>> 4d2dd8ae0a201b86a8ba480455a3a3966ad60b7b


```{r bray-cc}
bray_to_baseline_fltr %>% 
  filter(perturbation == "CC", RelDay >= -50, RelDay <= 50) %>%
  mutate(Interval = factor(Interval, level = names(cc_intv_cols))) %>%
  ggplot(
    aes(x = RelDay, y = dist_to_baseline, 
        group = Subject, color = Interval)) +
  geom_line(aes(group = Subject), alpha = 0.7, lwd = 0.5) + 
  geom_point(alpha = 0.5, size = 1.2) + 
  scale_color_manual(values = cc_intv_cols) + 
  theme(legend.position = "bottom") + 
  guides(colour = guide_legend(override.aes = list(size=3))) +
  xlab("Days from diet initiation") +
  ylab("Bray-Curtis distance to 7 pre-diet samples") 
```


```{r}
fpca.bray.cc30 <- fit_dist_to_baseline(
  bray_to_baseline_fltr %>% 
    filter(perturbation == "CC", RelDay >= -30, RelDay <= 30) %>%
    arrange(Subject, RelDay))

save(list = c("fpca.bray.abx", "fpca.bray.diet","fpca.bray.diet30", 
              "fpca.bray.cc", "fpca.bray.cc30"), 
     file = "output/fpca_res.rda")
```

```{r fpca-bray-cc, fig.width=12, fig.height=6.5}
(pCC <- bray_to_baseline_fltr %>% 
  filter(perturbation == "CC", RelDay >= -30, RelDay <= 30) %>%
ggplot(aes(x = RelDay, y = dist_to_baseline)) +
  geom_line(
    data = fpca.bray.cc30[["fitted"]],
    aes(group = Subject, x = time, y = value),
    alpha = 0.3, size = 0.7, color = "grey30") +
  geom_vline(xintercept = 0, lwd = 1, color = "orange") +
  geom_vline(xintercept = 4, lwd = 1, color = "orange") +
  geom_point(aes(color = Interval), size = 1.5, alpha = 0.7) +
  geom_line(
    data = fpca.bray.cc30[["mean"]], aes(x = time, y = value),
    color = "navy", size = 2) +
  scale_color_manual(values = cc_intv_cols, name = "Interval") +
  scale_x_continuous(
    name = "Days from initial antibiotic dose",
    limits = c(NA, NA), breaks = seq(-50, 60, 10)) +
  scale_y_continuous(
    name = "Bray-Curtis distance to baseline", 
    limits = c(NA, NA), breaks = seq(0.1, 0.80, 0.1)) +
  theme(text = element_text(size = 20))) 
```


```{r fpca-deriv-bray-cc, fig.width=12, fig.height=6.5}
fpca.bray.cc30[["fitted"]] <- fpca.bray.cc30[["fitted"]] %>%
  mutate(Interval = ifelse(fpca.bray.cc30[["fitted"]]$time < 0 , "PreCC","PostCC"))

(pCCDeriv <-  fpca.bray.cc30[["fitted"]] %>%
  ggplot(aes(x = RelDay)) +
  geom_line(
    aes(group = Subject, x = time, y = deriv, color = Interval),
    alpha = 0.5, size = 0.7) +
  geom_vline(xintercept = 0, lwd = 1, color = "orange") +
  geom_vline(xintercept = 4, lwd = 1, color = "orange") +
  scale_color_manual(values = cc_intv_cols, name = "Interval") +
  scale_x_continuous(name = "",
    limits = c(NA, NA), breaks = seq(-50, 60, 10)) +
  ylab("Derivative"))
```


```{r fpca-bray-cc-labs, fig.width=12, fig.height=6.5}
(pCCLab <- bray_to_baseline_fltr %>% 
  filter(perturbation == "CC", RelDay >= -50, RelDay <= 50) %>%
ggplot(aes(x = RelDay, y = dist_to_baseline)) +
  geom_line(
    data = fpca.bray.cc[["fitted"]],
    aes(group = Subject, x = time, y = value),
    alpha = 0.3, size = 0.7 ) +
  geom_vline(xintercept = 0, lwd = 1, color = "orange") +
  geom_vline(xintercept = 4, lwd = 1, color = "orange") +
  geom_text(
      aes(label = Subject, color = Interval), size = 4, alpha = 0.7) +
  geom_line(
    data = fpca.bray.cc[["mean"]], aes(x = time, y = value),
    color = "navy", size = 2) +
  scale_color_manual(values = cc_intv_cols, name = "Interval") +
  scale_x_continuous(
    name = "Days from initial antibiotic dose",
    limits = c(NA, NA), breaks = seq(-50, 60, 10)) +
  scale_y_continuous(
    name = "Bray-Curtis distance to baseline", 
    limits = c(0.1, 0.85), breaks = seq(0.1, 0.80, 0.1)) +
  theme(text = element_text(size = 20)))

```


```{r fpca-bray-cc-val-deriv, fig.width=12, fig.height=6.5}
vp = grid::viewport(width = 0.3, height = 0.32, x = 0.07, y =0.95, just = c("left", "top"))
print(pCC)
print(pCCDeriv + theme_bw(base_size = 10) + theme_subplot, vp = vp)
```

```{r}
sessionInfo()
```

