---
title: "Prediction"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(phyloseq)
library(RColorBrewer)
study_arm <- "Abx"
perturbation_relday <- "Abx_RelDay"
```

# Load data

```{r}
list.files("../../data/16S/phyloseq/")
```


```{r}
# File generated in /perturbation_16s/analysis/holmes_group_presentation/new_16S_analysis.rmd
ps <- readRDS("../../data/16S/phyloseq/perturb_physeq_fltr2430_17Jul18.rds")
ps
```

```{r}
SMP <- data.frame(ps@sam_data, stringsAsFactors = FALSE) %>%
  mutate(SampleDepth = sample_sums(ps))
  
SUBJ <-  SMP %>%
  select(Subject, Group, Age:BMI) %>%
  distinct()
```


```{r}
data.frame(ps@sam_data) %>%
    select(Group, Subject) %>%
    distinct() %>%
    group_by(Group) %>%
    summarise(subject_count = n())
```

## Generate subsets of subjects

```{r}
ps_subset <- subset_samples(ps, grepl(study_arm, sample_data(ps)$Group))
ps_subset
```

```{r}
# filter out dates immediately before and after the perturbation
ps_subset <- subset_samples(
    ps_subset, 
    sample_data(ps_subset)[[perturbation_relday]] >= -7 &
    sample_data(ps_subset)[[perturbation_relday]] <= 90) 
ps_subset
```

# Bray-curtis diversity

```{r}
# Compute bray-Curtis distance on asinh data
brayD.ihs <- phyloseq::distance(
    transform_sample_counts(ps_subset, function(x) {asinh(x)}), 
    method = "bray")
class(brayD.ihs)
```


```{r}
brayD.df <- reshape2::melt(
  as.matrix(brayD.ihs), 
  varnames = c("Sample_1", "Sample_2"),
  value.name = "bray") %>%
  mutate(Sample_1 = as.character(Sample_1),
         Sample_2 = as.character(Sample_2))
#dim  2152089       3

brayD.df <- brayD.df %>%
  left_join(
    SMP %>% select(Meas_ID, Subject), 
    by = c("Sample_1" = "Meas_ID"))

brayD.df <- brayD.df %>%
  left_join(
    SMP %>% select(Meas_ID, Subject), 
    by = c("Sample_2" = "Meas_ID"), 
    suffix = c("_1", "_2"))
 
brayD.df <- brayD.df %>%
    filter(Subject_1 == Subject_2) %>%
    rename(Subject = Subject_1) %>%
    select(-Subject_2)
head(brayD.df)
```


```{r}
brayD.df <- brayD.df %>%
    left_join(
        SMP %>% select(Meas_ID, Group, Abx_Interval, Abx_RelDay), 
        by = c("Sample_1" = "Meas_ID")) %>%
    left_join(
        SMP %>% select(Meas_ID, Group, Abx_Interval, Abx_RelDay), 
        by = c("Sample_2" = "Meas_ID"), suffix = c("_1", "_2")) 
dim(brayD.df)
```

Mean shift from baseline samples [-7, 0] measured with Bray-Curtis distance.

```{r}
brayD_mean_to_baseline <- brayD.df %>%
    filter(Abx_RelDay_1 >= -7 & Abx_RelDay_1 <= 0) %>%
    group_by(Subject, Group_2, Abx_Interval_2, Abx_RelDay_2) %>%
    summarize(mean_bray = mean(bray))

brayD_mean_to_baseline <- brayD_mean_to_baseline %>%
    rename(Group = Group_2, Abx_Interval = Abx_Interval_2, 
           Abx_RelDay = Abx_RelDay_2)
dim(brayD_mean_to_baseline)  
```

```{r, fig.width=8}
brayD_mean_to_baseline %>%
ggplot(
    aes_string(x = paste0(study_arm, "_RelDay"), y = "mean_bray", 
               group = "Subject", color = paste0(study_arm, "_Interval"))) +
    geom_point() + geom_line(aes(group = Subject)) +
    theme(legend.position = "bottom")
```



## Compute response measures

First find the time to stabilization

```{r}
brayD_mean_to_baseline <- brayD_mean_to_baseline %>%
    group_by(Subject) %>%
    mutate(
        mean_post = mean(mean_bray[Abx_RelDay >= 20 & mean_bray < 0.6]), # guard against outliers
        sd_post = sqrt(var(mean_bray[Abx_RelDay >=20 & mean_bray < 0.6]))
    ) 
tolerance <- brayD_mean_to_baseline %>% select(Subject, sd_post) %>% 
    distinct() %>% .[["sd_post"]] %>% mean()

brayD_mean_to_baseline <- brayD_mean_to_baseline %>%
    mutate(
        within_sd = Abx_RelDay > 5 & abs(mean_bray - mean_post) < tolerance) %>%
    arrange(Subject, Abx_RelDay) %>%
    group_by(Subject) %>%
    mutate(stab = cumsum(within_sd) >= 2 )
```

```{r}
time_to_stab <- brayD_mean_to_baseline %>% 
    group_by(Subject) %>%
    filter(stab) %>%
    top_n(n = 1, wt = -Abx_RelDay) %>%
    rename(time_to_stab = Abx_RelDay)

subj_set <- time_to_stab %>% filter(time_to_stab >= 30)
```


```{r, fig.width=10, fig.height=8}
brayD_mean_to_baseline %>%
    filter(Abx_RelDay > -7, Abx_RelDay < 100) %>%
   # filter(Subject %in% sbj) %>%
   # filter(Subject == bray_recovery$Subject[bray_recovery$slope_pert > 0.6]) %>%
ggplot(
    aes_string(x = paste0(study_arm, "_RelDay"), y = "mean_bray", 
               group = "Subject", color = "stab")) +
    geom_point() +
    geom_text(aes(label = Subject)) + 
    geom_line(aes(group = Subject)) +
    theme(legend.position = "bottom")
```


```{r}
# CHANGE HERE for other study arms
bray_recovery <- brayD_mean_to_baseline %>%
    group_by(Subject) %>%
    summarize(
        time_to_stab = min(Abx_RelDay[stab]),
        max_amplitude = max(mean_bray[Abx_RelDay >= 0 & Abx_RelDay <= 10]),
        day_max_amplitude = min(Abx_RelDay[mean_bray == max_amplitude]),
        deviation_MidAbx = mean(mean_bray[Abx_RelDay %in% c(3,4,5)]),
        deviation_Stab = mean_bray[Abx_RelDay == time_to_stab],
        slope_t4 =  mean(mean_bray[which.min(abs(Abx_RelDay-4))+1] - 
            mean_bray[which.min(abs(Abx_RelDay-4))]),
        slope_t8 =  mean(mean_bray[which.min(abs(Abx_RelDay-8))+1] - 
            mean_bray[which.min(abs(Abx_RelDay-8))]),
        slope_t14 =  mean(mean_bray[which.min(abs(Abx_RelDay-14))+1] - 
            mean_bray[which.min(abs(Abx_RelDay-14))]),
        slope_pert = max_amplitude/day_max_amplitude,
        slope_return = mean((deviation_Stab - max_amplitude)/time_to_stab)
    ) 
```

```{r}
bray_recovery
```


## Measures for pre-perturbation variability

```{r}
daily_variability <- brayD.df %>%
    mutate(day_diff = Abx_RelDay_2 - Abx_RelDay_1) %>%
    filter(day_diff <= 2, day_diff > 0, Abx_RelDay_2 <= 0) 
    
daily_variability_mean <- daily_variability %>%
    group_by(Subject) %>%
    summarise(preAbx_variability = mean(bray), n_for_var = n()) 
head(daily_variability_mean)
```


```{r}
smp_info <- data.frame(
    sample_data(ps) %>% 
        select(Meas_ID, Samp_ID, Subject, contains("Abx_"), 
               Group, DaysFromStart, Samp_Date),  
    richness = apply(as(otu_table(ps), "matrix"), 2, function(x) sum(x > 0))
)

subj_data <- smp_info %>% 
    filter(grepl("Abx", Group), Abx_RelDay >= -7, Abx_RelDay <= 0) %>%
    group_by(Subject, Group) %>% 
    summarise(preAbx_richness = mean(richness), n_for_rich = n()) %>%
    ungroup() %>%
    left_join(daily_variability_mean)
```


```{r fig.width=12, fig.height=12, message=FALSE}
library(GGally)
df <- subj_data %>% left_join(bray_recovery)
plt <- df %>% 
    filter(!Subject %in% c("BAA", "BAN")) %>%
    ggpairs(columns = c(3, 5, 7, 8, 9, 12:15))
plt
```

```{r}
bray_recovery %>% filter( slope_pert > 0.6)
```


```{r, fig.width=10, fig.height=8}
brayD_mean_to_baseline %>%
   filter(Subject == bray_recovery$Subject[bray_recovery$slope_pert > 0.6]) %>%
ggplot(
    aes_string(x = paste0(study_arm, "_RelDay"), y = "mean_bray", 
               group = "Subject", color = "stab")) +
    geom_point() +
    geom_text(aes(label = Subject)) + 
    geom_line(aes(group = Subject)) +
    theme(legend.position = "bottom")
```



## Fitting functional pca

```{r}
brayD_mean_to_baseline %>% 
    filter(Abx_RelDay > 80) %>%
    with(table(Subject))
```

We will use only data from time < 80 to fit ecause not many subjects have 
samples beyond.

```{r}
library("fdapace")
source('./fda_pca_funs.R')

bray_fpca <- fit_fdapca(
  brayD_mean_to_baseline %>% 
      filter(Abx_RelDay <= 80) %>%
      arrange(Subject, Abx_RelDay), 
  time_column = "Abx_RelDay",
  value_column = "mean_bray", 
  replicate_column = "Subject",
  cluster = FALSE, filter = FALSE, 
  fpca_optns = list(nRegGrid = 88))

bray_mean <- data.frame(
  time = bray_fpca$workGrid,
  value = bray_fpca$mu)

bray_fit <- fpca_fit(bray_fpca) %>%
    rename(Subject = Replicate_ID) 

bray_fit <- bray_fit %>%
    left_join(
        fpca_fit(bray_fpca, derOptns = list(p = 1)) %>%
            rename(Subject = Replicate_ID) %>%
            rename("deriv" = value)
    )
```

```{r, fig.width=10, fig.height=8}
 plt <-  bray_fit %>%
    ggplot(aes(x = time, y = value)) +
        geom_line(aes(group = Subject), alpha = 0.3, size = 0.7 ) +
        geom_line(
            data = bray_mean,
            aes(x = time, y = value),
            color = "navy", size = 2.5) +
        geom_text(
            size = 3,
            data = brayD_mean_to_baseline,
            aes_string(
                x = paste0(study_arm, "_RelDay"), y = "mean_bray",
                label = "Subject", color = paste0(study_arm, "_Interval"))) +
        geom_vline(xintercept = 0, lwd = 1, color = "orange") +
        geom_vline(xintercept = 4.5, lwd = 1, color = "orange") +
        theme(legend.position = "bottom") 
print(plt)
```


```{r abx-bray-fit, fig.width=8, fig.height=5}
for (subj in unique(brayD_mean_to_baseline$Subject)) {
   plt <-  bray_fit %>%
    filter(Subject %in% subj)%>%
    ggplot(aes(x = time, y = value)) +
        geom_line(aes(group = Subject), alpha = 0.3, size = 0.7 ) +
        geom_line(
            data = bray_mean,
            aes(x = time, y = value),
            color = "navy", size = 2.5) +
        geom_text(
            size = 3,
            data = brayD_mean_to_baseline %>%  
                filter(Abx_RelDay <= 60) %>%
                filter(Subject %in% subj),
            aes_string(
                x = paste0(study_arm, "_RelDay"), y = "mean_bray",
                label = "Subject", color = paste0(study_arm, "_Interval"))) +
        geom_vline(xintercept = 0, lwd = 1, color = "orange") +
        geom_vline(xintercept = 4.5, lwd = 1, color = "orange") +
        theme(legend.position = "bottom") 
   print(plt)
}

```


```{r}
nthresh <- 3
stab_tol <- 1e-3

subj_fpca_est <- bray_fit %>%
    arrange(Subject, time) %>%
    group_by(Subject) %>%
    mutate(
        sum_stab = cumsum(ifelse(time >= 5, abs(deriv) < stab_tol, FALSE)),
        stab = sum_stab > nthresh
)

subj_fpca_aggr <- subj_fpca_est %>%
    group_by(Subject) %>%
    summarise(
        time_to_stab_fpca = min(time[stab]),
        deriv_t4 = deriv[which.min(abs(time - 4))],
        deriv_t6 = deriv[which.min(abs(time - 6))],
        deriv_t8 = deriv[which.min(abs(time - 8))],
        deriv_t10 = deriv[which.min(abs(time - 10))],
        deriv_t12 = deriv[which.min(abs(time - 12))],
        deriv_t14 = deriv[which.min(abs(time - 14))],
        deriv_t16 = deriv[which.min(abs(time - 16))]
    )
```


```{r, fig.width=6, fig.height=4}
 plt <-  bray_fit %>%
    left_join(subj_fpca_est) %>%
    ggplot(aes(x = time, y = value)) +
     geom_text(
            size = 3,
            data = brayD_mean_to_baseline %>%
                left_join(subj_fpca_est %>% select(Subject, time, stab),
                          by=c("Subject", "Abx_RelDay"="time")),
            aes_string(
                x = paste0(study_arm, "_RelDay"), y = "mean_bray",
                color = "stab.y", label = "Subject")) +
        geom_line(
            aes(group = Subject, color = stab), size = 0.7 ) +
        geom_line(
            data = bray_mean,
            aes(x = time, y = value),
            color = "navy", size = 2.5) +
   
        geom_vline(xintercept = 0, lwd = 1, color = "orange") +
        geom_vline(xintercept = 4.5, lwd = 1, color = "orange") +
        theme(legend.position = "bottom") 
print(plt)
```


```{r, fig.width=10, fig.height=8}
 plt <-  bray_fit %>%
    left_join(subj_fpca_est) %>%
    ggplot(aes(x = time, y = deriv)) +
        geom_line(
            aes(group = Subject, color = stab), size = 0.7 ) +
        geom_vline(xintercept = 0, lwd = 1, color = "orange") +
        geom_vline(xintercept = 4.5, lwd = 1, color = "orange") +
        theme(legend.position = "bottom") 
print(plt)
```




```{r fig.width=12, fig.height=12, message=FALSE}
recovery_df <- subj_data %>% 
  left_join(bray_recovery) %>% left_join(subj_fpca_aggr)


plt <- recovery_df %>% 
    filter(!Subject %in% c("BAA", "BAN")) %>%
    ggpairs(columns = c(3, 5, 9, 7, 17, 8, 12, 14, 18, 23))
plt
```



```{r fig.width=9.5, fig.height=9.5, message=FALSE}
recovery_df <- subj_data %>% 
  left_join(bray_recovery) %>% left_join(subj_fpca_aggr)


plt <- recovery_df %>% 
    filter(!Subject %in% c("BAA", "BAN")) %>%
    ggpairs(columns = c(3, 5, 9, 7, 8, 18, 23))
plt + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
counts <- as(otu_table(ps), "matrix")
counts[1:5, 1:5]
```


```{r}
qplot(log(sample_sums(ps)))
```


```{r}
# library(factoextra)
# pca_res <- prcomp(t(asinh(counts)))
# 
# fviz_pca_ind(pca_res)
```




```{r}
sessionInfo()
```

